---
title: "proteomics_preprocessing"
output: html_document
date: "2026-08-11"
---

## Data source & preprocessing notes

Raw data: `pr_00086_hong_20250819_Proteins.xlsx`, TMTpro-labeled LC-MS/MS 
(Orbitrap Fusion Lumos, MS3/SPS quantification), 12 FFPE epilepsy surgery
samples (6 good / 6 poor outcome), all run in a single TMT multiplex.

**Proteins are pre-filtered by the core to those detected in all 12
samples** — no missing values by design.

**Two abundance types are exported; we use "Abundance," not "IBAQ."**
"Abundance" = raw TMT reporter-ion intensities, normalized by the core
to total signal per channel — valid for comparing a given protein across
samples/outcome groups. "IBAQ" = those same values further normalized by 
protein molecular weight, intended for comparing *different* proteins to 
each other within a sample.


```{r}
library(dplyr)
library(DEqMS)
```

# Read in data & filter
```{r}
datapath <- "~/Documents/repos/epimap/data"
raw <- readxl::read_excel(paste0(datapath, "/protein/pr_00086_hong_20250819_Proteins.xlsx"), 
                          sheet = "protein abundances")
meta = read.csv(paste0(datapath, "/clinical_metadata_260811.csv"))

id_mapping <- meta$patient_id
names(id_mapping) <- sub("^", "Abundance ", meta$proteomics_id)

# rename column names
idx <- match(names(id_mapping), colnames(raw))
colnames(raw)[idx] <- id_mapping[!is.na(idx)]

# filter out duplicates (pick one with more peptides)
raw_filt <- raw %>%
  filter(`Gene Symbol` != "" & !is.na(`Gene Symbol`)) %>%
  group_by(`Gene Symbol`) %>%
  slice_max(`Number of Unique Peptides`, n = 1, with_ties = FALSE) %>%
  ungroup()

print(paste0("Number of proteins after removing duplicates: ", nrow(raw_filt)))

# filter out proteins mapped by fewer than 2 peptides
raw_filt <- raw_filt %>% filter(`Number of Unique Peptides` >= 2)
print(paste0("Number of proteins after removing those mapped by fewer than 2 peptides: ", nrow(raw_filt)))
```


# Normalise and transform
```{r}
raw_filt <- raw_filt %>%
  select(`Gene Symbol`, contains("MTS"))

# plot distribution: frequency vs sample intensities
raw_melt <- raw_filt %>% tidyr::pivot_longer(-`Gene Symbol`, names_to = "sample", values_to = "value")
hist(raw_melt$value, breaks = 100, 
     xlab = "IBAQ intensity", ylab = "Frequency", main = "Freq vs IBAQ intensity (before log2)")

# log2 transform
raw_log2t <- log2(tibble::column_to_rownames(raw_filt, "Gene Symbol"))

# plot distribution: frequency vs sample intensities
raw_log2t_melt <- raw_log2t %>% melt()
hist(raw_log2t_melt$value, breaks = 100, 
     xlab = "IBAQ intensity", ylab = "Frequency", main = "Freq vs IBAQ intensity (after log2)")

# median normalisation
raw_log2t_norm <- equalMedianNormalization(raw_log2t)
raw_log2t_norm_melt <- raw_log2t_norm %>% melt()
hist(raw_log2t_norm_melt$value, breaks = 100, 
     xlab = "IBAQ intensity", ylab = "Frequency", main = "Freq vs IBAQ intensity (after log2 & normalisation)")
```

Histograms look very similar pre and post median normalisation.

```{r}
processed <- tibble::rownames_to_column(raw_log2t_norm, "gene_name")
processed %>% left_join(., raw[])

write.csv(processed, "~/Documents/repos/epimap/data/protein/processed_protein_260811.csv", row.names = FALSE)
```

